Supplemental Material for: The evolution of gene expression in the term placenta of viviparous mammals

### 2017-02-15 manuscript/manuscript.Rnw
bold core functionality
discuss orthologies by species type

To address our questions, we 
sequenced and quantified the total RNA of mammalian placentas at term,
including \human, \bonobo (the sister
species to \textit{Pan troglodytes}), \smonkey,
\mouse, \galili, \carmeli, \dog, \cow, \elephant,
\armadillo, and \opossum.
These species represent four major branches of eutherian mammals
(Euarchontoglires, Laurasiatheria, Afrotheria, and Xenartha) as well
as marsupials. These placentas also represent a wide array of
placental morphologies as well as traits such as number and size of
offspring and length of gestation (\cref{tab:gestational_periods}). We
use these data, along with existing RNA-Seq data from \mouse
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
\horse
\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and
\pig \cite{Ahanda.ea2012:Predictionofaltered3-} to: 1)
identify the non-housekeeping genes which are expressed in the
placentas of all eutherian species analyzed and therefore represent the
\emph{core functionality of the placenta}; 2) identify genes whose
expression level changes significantly in each of the morphology
types.

@@

Single-copy (1-to-1) orthologs in were determined using
Ensembl v.80 gene trees via Biomart using the biomaRt R package
\cite{Durinck.ea2009:Mappingidentifiersforintegration}. %
A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} single-copy orthologous genes were expressed
in at least one species.

@@

Supplemental \cref{fig:transcriptome_statistics} shows the RNA-Seq read and
alignment statistics for all species. For transcriptomes sequenced
with 76 bp %
%
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])}),%
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (Supplemental \cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 100 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 80 & input.read.length < 120,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 80 & input.read.length < 120,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 80 & input.read.length < 120,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 80 & input.read.length < 120,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (Supplemental \cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 35 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with an average of %
\Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
%
of reads mapping to the respective reference genome (Supplemental
\cref{tab:ref_genome_aligned_species}). Read length was not a
significant predictor of unmapped read percentage (Supplemental
\cref{fig:transcriptome_reads_vs_read_length}). The expression of
all genes with \ac{FPKM} $\ge 50$ is shown in Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.

>>>>

To address our questions, we 
sequenced and quantified the total RNA of mammalian placentas at term,
including \human, \bonobo (the sister
species to \textit{Pan troglodytes}), \smonkey,
\mouse, \galili, \carmeli, \dog, \cow, \elephant,
\armadillo, and \opossum.
These species represent four major branches of eutherian mammals
(Euarchontoglires, Laurasiatheria, Afrotheria, and Xenartha) as well
as marsupials. These placentas also represent a wide array of
placental morphologies as well as traits such as number and size of
offspring and length of gestation (\cref{tab:gestational_periods}). We
use these data, along with existing RNA-Seq data from \mouse
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
\horse
\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and
\pig \cite{Ahanda.ea2012:Predictionofaltered3-} to: 1)
identify the non-housekeeping genes which are expressed in the
placentas of all eutherian species analyzed and therefore represent the
\textbf{core functionality of the placenta}; 2) identify genes whose
expression level changes significantly in each of the morphology
types.

@@

Single-copy (1-to-1) orthologs in were determined using Ensembl v.80
gene trees via Biomart using the biomaRt R package
\cite{Durinck.ea2009:Mappingidentifiersforintegration}. %
A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} single-copy orthologous genes were
expressed in at least one species. We also obtained orthology from the
perspective of each species to all other species using biomaRt package
and Ensembl v.80 database to determine the percentage of highly
expressed genes which had an orthology of one-to-one, one-to-many,
many-to-many, or novel/unknown.

@@

Supplemental \cref{fig:transcriptome_statistics} shows the RNA-Seq read and
alignment statistics for all species. For transcriptomes sequenced
with 76 bp %
%
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])}),%
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (Supplemental \cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 100 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 80 & input.read.length < 120,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 80 & input.read.length < 120,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 80 & input.read.length < 120,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 80 & input.read.length < 120,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (Supplemental \cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 35 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with an average of %
\Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
%
of reads mapping to the respective reference genome (Supplemental
\cref{tab:ref_genome_aligned_species}). Read length was not a
significant predictor of unmapped read percentage (Supplemental
\cref{fig:transcriptome_reads_vs_read_length}). %
%
Between 20\% and 40\% of all highly expressed genes in each species
were single-copy orthologs with all other analyzed species, with the
remainder being mostly one-to-many or many-to-many in comparison to at
least one other species (Supplemental
\cref{tab:orthologs_in_each_species}).
%
The expression of
all genes with \ac{FPKM} $\ge 50$ is shown in Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.

### 2017-02-15 manuscript/supplemental_material.Rnw
discuss TSI in manuscript; move enrichment table
discuss orthologies by species type

>>>>

\subsection{Orthology types in each Species}

\begin{table}[h]
  \footnotesize
<<orthologs_in_each_species,echo=FALSE,results="asis">>=
load("../data/orthologs_by_species_ranking")
orthologs.by.species[,percentage:=ratio*100]
orthologs.by.species[,orthology_type:=
                          factor(orthology_type,
                                 levels=c("one2one",
                                          "one2many",
                                          "many2many",
                                          "novel/unknown"))]
orthology.wide.table <-
    data.table(dcast(orthologs.by.species,
                     type+orthology_type~species,
                     value.var="percentage"))
setnames(orthology.wide.table,
         "type","Cutoff")
orthology.wide.table[,Cutoff:=xtable.sanitize.text(Cutoff)]
orthology.wide.table[,orthology_type:=xtable.sanitize.text(orthology_type)]
setnames(orthology.wide.table,
         "orthology_type","Orthology Type")
setnames(orthology.wide.table,
         colnames(orthology.wide.table)[-(1:2)],
         paste0("\\textit{",
                gsub("^(.)","\\U\\1",
                     colnames(orthology.wide.table)[-(1:2)],
                     perl=TRUE),
                "}"))
print(xtable(orthology.wide.table,
             digits=0
             ),
      rotate.colnames=TRUE,
      booktabs=TRUE,
      floating=FALSE,
      include.rownames=FALSE,
      sanitize.text.function=genes.table.sanitize
      )
@ 
\caption{Percentage of orthology in the top 100 and top 1\% of
  placenta transcripts as identified by Ensembl. If a gene has the
  same orthology in comparison to all sequenced species, it is
  reported as that orthology; if it had an orthology of many2many in
  any species, it is reported as many2many; if it had an orthology of
  one2many in any species, but no many2many, it is reported as
  one2many. Novel/unknown genes had no reported orthologies to any
  sequenced species.}
\label{tab:orthologs_in_each_species}
\end{table}
